A Simple Minibatch Acceptance Test for MCMC

### 2016-05-18 nips_2016.tex
Made major changes to Section 3 to clarify things but still in a draft stage. Also added my python notebook to generate Figure 1.

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} propose an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on the likelihood ratio. If the test cannot make a decision
over a certain confidence threshold, then they increase the minibatch size and test again. This
process repeats until a decision.  The downside of this algorithm is the need to keep incrementing
the mini-batch size in one iteration; in the worst case, all $N$ data points may be needed in a
single iteration.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{mh_convolution_diagram}
  \caption{
  {\color{blue}
  Daniel: this image is finalized, but change this caption.
  }
  }
  \label{fig:deconvolution}
\end{figure}

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\beta \Delta))^{-1}$. Again, straightforward arithmetic shows that it satisfies the
condition in Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily
sample from it using the following procedure. At any iteration with a current parameter $\theta_t$
and a candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u
\sim {\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. To see why, define $X = g^{-1}(u)$ where
again, $g$ is the logistic function. The CDF of $X$ is precisely the logistic function because for
arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of the uniform random variable here is simply one. Thus, the criteria to accept the
candidate $\theta'$ is equivalent to whether $\Delta > X$. Since $X$ is symmetric about zero, the
acceptance criteria is equivalently expressed as $\Delta + X>0$.

@@

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly, then the distribution of $\Delta'$ is approximately Gaussian.
\end{lemma}

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian by the
Central Limit Theorem.
\end{proof}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship can be expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise in $\Delta'$ means we can no longer perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. We can decompose
$X$ as $X = X_{\rm norm}+X_{\rm corr}$, where the first term is a zero-mean ``noise term'' which
has the same variance as $\varepsilon$, and the second term is a zero-mean ``correction'' term.
These two convolve to form $X$. The criteria to accept can be rewritten as
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
This deconvolution process is visualized in Figure~\ref{fig:deconvolution}. We decompose $X$ this
way so that we can estimate $X_{\rm norm}$ from the data, and then add the correction $X_{\rm corr}$
later.  Notice that we have some control over the true value of $X_{\rm norm}$, even though we
cannot generally guarantee computing it exactly. In particular, increasing the minibatch size will
cause the $X_{\rm norm}$ distribution to become narrower. {\color{blue} (Daniel: note that we will
be fixing this figure, so the visualization should hopefully be a little clearer, along with the
corresponding paragraph description. In particular, this paragraph needs an informal description of
Algorithm~\ref{alg:our_algorithm}.)} Algorithm~\ref{alg:our_algorithm} concisely describes our MH
test.

Since our proposed acceptance function $g(\Delta) = (1+ \exp(-\beta \Delta))^{-1}$ satisfies
detailed balance (see Lemma~\ref{lem:detailed_balance}), if $\Delta$ is computed from the complete
data, the distribution of the sampled data will match our target distribution $f$, the posterior,
under standard MCMC assumptions (e.g., that the chain is run sufficiently long enough and mixes
well). Now consider the case of using minibatches of data. Let our true and estimated acceptance
probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} >
0)$.  If our estimate of $X_{\rm norm}$ is exact, then $P_a' = P_a$, which follows by making the
inequality in Equation~\ref{eq:criteria} an equality.

It is also important to note that our minibatch MH test can also be used to obtain parameter
estimates close to the MAP and/or ML estimaes by augmenting the likelihood function with a
temperature term $T > 1$. Thus, the target distribution would become
\begin{equation}
\label{eq:temperature} f(\theta') = p(\theta')\left ( \prod_{i=1}^N p(x_i \mid \theta')\right)^{\frac{1}{T}}.
\end{equation}
We can decrease the temperature as the MCMC sampling proceeds, thus slowing the mixing rate and
sharpening the peaks of the distribution, so that the samples coverge to the desired point estimate.

{\color{blue}
Daniel: we really need to make this clear and do the algorithm description.
}

\begin{algorithm}[t]
\KwData{$\{x_i\}$ under a target distribution}
\KwResult{a serise of sampled $\{\theta_i\}$ under the posterior distribution} 
set the $var(\Delta') = 1.0$ and deconvolve $X$ to get the disribution of $X_{corr}$\;
\While{not enough sampled data}{
	estimate the $sd(\Delta')$\;
    propose a candidate $\theta'$\;
    compute the noisy $\Delta'$\;
    sample a $x$ from $X_{corr}$\;
    \eIf{$x*sd(\Delta') + \Delta' > 0$}{
        Accept the $\theta'$;
    }{
        Reject the $\theta'$;
    }
}
\caption{A description of our MH test.}
\label{alg:our_algorithm}
\end{algorithm}

>>>>

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on the likelihood ratio. If the test cannot make a decision
over a certain confidence threshold, then they increase the minibatch size and test again. This
process repeats until a decision.  The downside of this algorithm is the need to keep incrementing
the mini-batch size in one iteration; in the worst case, all $N$ data points may be needed in a
single iteration.

@@

@@

For our MH test, the key is that we use a different $g$, the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Again, straightforward arithmetic shows that it satisfies the
condition in Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily
sample from it using the following procedure. At any iteration with a current parameter $\theta_t$
and a candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u
\sim {\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. We can define $X = g^{-1}(u)$ so that its CDF
is the logistic function, which follows because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, $X$ is symmetric about zero, so the acceptance criteria can also be
expressed as $\Delta + X>0$.

@@

\begin{lemma}\label{lem:gaussian}
If the minibatch data is chosen randomly and sufficiently large, then the distribution of $\Delta'$ is approximately Gaussian.
\end{lemma}

\begin{proof}
By expanding the definition of $\Delta'$ from Equation~\ref{eq:deltas}, we get
\[
\Delta' = \log p(\theta') - \log p(\theta_t) + \log q(\theta_t \mid \theta') - \log q(\theta' \mid \theta_t) +
\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t)).
\]
During a given iteration, $\theta_t$ and $\theta'$ are fixed. Therefore, the randomness in $\Delta'$
comes only from the minibatch. Since the minibatch is chosen randomly, the distribution of
$\sum_{i=1}^n (\log p(x_i\mid \theta') - \log p(x_i\mid \theta_t))$ converges to a Gaussian as $n
\to \infty$ by the Central Limit Theorem.
\end{proof}

% Daniel: I decided to put the figure here, so it's closer to the text that references it.
\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{mh_convolution_diagram}
    \caption{
    Three examples of $X_{\rm norm}$ and $X_{\rm corr}$ distributions that convolve to form the
    standard logistic distribution ($\mu=0$ and $s=1$). The three distributions correspond to three
    different standard deviation (SD) values of the $X_{\rm norm}$ distribution (the colors carry
    over to the $X_{\rm corr}$ plot, whose SD values correspond to the $X_{\rm norm}$ SD values).
    This figure must be viewed in color.
    }
    \label{fig:deconvolution}
    \vspace{-10pt}
\end{figure}

From Lemma~\ref{lem:gaussian}, since $\Delta'$ is a noisy approximation of $\Delta$, the
relationship is expressed as
\begin{equation}\label{eq:relationship}
\Delta' = \Delta + \varepsilon, \quad \varepsilon \sim \mathcal{N}(0, \sigma^2),
\end{equation}
for a Gaussian noise term $\varepsilon$.  The extra noise means we can no longer directly perform
the $\Delta + X$ test, so we need to slightly change our acceptance criteria. We can decompose $X$
as
\begin{equation}\label{eq:deconvolution}
X = X_{\rm norm}+X_{\rm corr},
\end{equation}
where the first term has the same distribution as $\varepsilon$ and the second term is a zero-mean
``correction'' term.  These two add (i.e., convolve) to form $X$. The criteria to accept can be rewritten as
\begin{equation}\label{eq:criteria}
\Delta + X = \Delta + X_{\rm norm} + X_{\rm corr} \approx \Delta' + X_{\rm corr} >0.
\end{equation}
The approximation exists in Equation~\ref{eq:criteria} because, while we know the complete
distribution of $X$, the $X_{\rm norm}$ distribution must be estimated from minibatches of data.
Since $X_{\rm norm}$ is known a priori to be a zero-mean Gaussian, this task reduces to estimating
its standard deviation.

It is important to understand why we use Equations~\ref{eq:deconvolution} and~\ref{eq:criteria}. The
deconvolution process allows us to use $\Delta' + X_{\rm corr}$ as a proxy for $\Delta + X$. It
works because if we know the distribution of $\varepsilon$ (from Equation~\ref{eq:relationship},
which again reduces to knowing its standard deviation), then we can pretend that our $\Delta'$ is in
fact $\Delta$. To ``insert'' $\varepsilon$ into our test, we deconvolve $X$ so that one of its
components, $X_{\rm norm}$, plays the part of $\varepsilon$. This raises two immediate tasks: during
a given iteration, with current sample $\theta_t$ and proposed sample $\theta'$, how should we get
the densities of $X_{\rm norm}$ and $X_{\rm corr}$?

\begin{enumerate}[noitemsep]
    \item To estimate the variance of $X_{\rm norm}$, we generate $K$ values of $\Delta'$, each
    using a different random minibatch of $n$ data points (but each using the same $\theta_t$ and
    $\theta'$ parameters). In our experiments (see Section~\ref{sec:experiments}), $K$ is usually
    small, around five, and $n$ is usually a few hundred for the Central Limit Theorem to apply.
    Using these $K$ values, we estimate the empirical standard deviation ${\rm std}(\Delta')$.

    \item Once we have estimated ${\rm std}(\Delta')$, we can determine the distribution of
    $X_{\rm corr}$ from standard deconvolution techniques. These rely on the well-known fact that
    the Fourier transform of a convolution is the product of the Fourier transforms. For our
    experiments, we pre-compute $X_{\rm corr}$ values and store them in matrices so that during MCMC
    runs, we sample a standard uniform random variable and use that to directly obtain $X_{\rm
    corr}$.
\end{enumerate}

Figure~\ref{fig:deconvolution} demonstrates this deconvolution process and provides three examples
of densities for $X_{\rm norm}$ and $X_{\rm corr}$. We know that the densities of $X_{\rm norm}$ and
$X_{\rm corr}$ convolve to form the density of a standard logistic random variable with mean $\mu=0$
and scale $s=1$. If $X_{\rm norm}$ has small standard deviation, then the deconvolution to form
$X_{\rm corr}$ will result in a wider distribution. Likewise, with a larger standard deviation for
$X_{\rm norm}$, the $X_{\rm corr}$ distribution is narrower, but note that it becomes increasingly
unstable with bumps in the distribution, because the logistic curve has fatter tails than the
Gaussian curve. For this reason, we cap the standard deviation possibilities of $X_{\rm norm}$ at
$\approx 1.2$. We also set the minimum standard deviation of $X_{\rm norm}$ to be 0.1, because below
that, the $X_{\rm corr}$ distribution essentially matches the distribution of the logistic.

Algorithm~\ref{alg:our_algorithm} concisely describes our MH test.

Since our proposed acceptance function $g(\Delta) = (1+ \exp(-\beta \Delta))^{-1}$ satisfies
detailed balance (see Lemma~\ref{lem:detailed_balance}), if $\Delta$ is computed from the complete
data, the distribution of the sampled data will match our target distribution $f$, the posterior,
under standard MCMC assumptions (e.g., that the chain is run sufficiently long enough and mixes
well). Now consider the case of using minibatches of data. Let our true and estimated acceptance
probabilities be $P_a = g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} >
0)$.  If our estimate of $X_{\rm norm}$ is exact, then $P_a' = P_a$, which follows by making the
approximation in Equation~\ref{eq:criteria} an equality.

It is also important to note that our minibatch MH test can also be used to obtain parameter
estimates close to the MAP and/or ML estimates by augmenting the likelihood function with a
temperature term $T > 1$. Thus, the target distribution would become
\begin{equation}
\label{eq:temperature} f(\theta') = p(\theta')\left ( \prod_{i=1}^N p(x_i \mid \theta')\right)^{\frac{1}{T}}.
\end{equation}
We can decrease the temperature as the MCMC sampling proceeds, thus slowing the mixing rate and
sharpening the peaks of the distribution, so that the samples converge to the desired point estimate.

\begin{algorithm}[t]
\KwData{$\{x_i\}$ under a target distribution}
\KwResult{A series of sampled $\{\theta_i\}$ under the posterior distribution} 
set the $var(\Delta') = 1.0$ and deconvolve $X$ to get the distribution of $X_{corr}$\;
\While{not enough sampled data}{
	estimate the $sd(\Delta')$\;
    propose a candidate $\theta'$\;
    compute the noisy $\Delta'$\;
    sample a $x$ from $X_{corr}$\;
    \eIf{$x*sd(\Delta') + \Delta' > 0$}{
        Accept the $\theta'$;
    }{
        Reject the $\theta'$;
    }
}
\caption{A description of our MH test.}
\label{alg:our_algorithm}
\end{algorithm}

{\color{blue}
Daniel: lots of things to do for this section, such as revising the Algorithm 1 description,
breaking this up into subsections (it's a bit long), and also just making sure it is clear for
readers to understand.
}
